Floral preferences of mountain bumble bees are constrained by trait matching but flexible through elevation and season

### 2021-08-25 ms_2/ms_2_submission.Rmd
preped ms for Ecol Lett

---
title: "Floral preferences of mountain bumble bees are driven by trait matching but labile through elevation and time"

author: Douglas B. Sponsler^1^*, Katharina Kallnik^1^, Fabrice Requier^2^, Alice Classen^1^, A. Fabienne Maihoff^1^, Johanna Sieger^3^, Ingolf Steffan-Dewenter^1^

# date: `r Sys.Date()`

output:
  bookdown::word_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_md: yes
    number_sections: no
    toc: no
    reference_docx: Word_template.docx

bibliography: 
  - references.bib
#  - knitcitations.bib
csl: ecology.csl
# csl: https://www.zotero.org/styles/ecology

---

^1^ Department of Animal Ecology and Tropical Biology, Biocenter, University of Würzburg, Würzburg, Germany

^2^ Université Paris-Saclay, CNRS, IRD, UMR Évolution, Génomes, Comportement et Écologie, 91198, Gif-sur-Yvette, France

^3^ Naturpark Frankenhöhe, Ansbach, Germany

`*` Corresponding author: douglas.sponsler@uni-wuerzburg.de

@@

Patterns of resource use by animals can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Using ~13,000 records of bumble- bee-wildflower interactions along a 1400 m elevation gradient, we find that the selection of floral morphotypes by bumble bees is driven both by the directed process of trait-matching and the neutral effects of floral abundance. Patterns of floral preference are, moreover, labile through time and elevation, which may be an adaptation to the extreme spatiotemporal floristic turnover of mountain ecosystems. The flexibility of bumble bee foraging will be tested in coming years as climate and land use change drive rapid changes in floral resource distribution in time and space, particularly in the mountain ecosystems that host the world's richest communities of bumble bees and their floral mutualists.

*Keywords*: foraging, trait matching, pollinator, *Bombus*, alpine, niche, resource selection

@@

Patterns of resource selection and partitioning by animals are instrumental in community assembly, species coexistence, and ecosystem function [@Chesson2000-aw; @Bascompte2007-ls]. Resource selection is fundamentally constrained by phenotypic compatibility (i.e. trait-matching) between resources and consumers, but it can also be influenced by the relative availability of resources [@Johnson1980-co] and modulated by exogenous drivers such as competition [@Pianka1974-qk]. Understanding these interacting processes is key to anticipating the response of ecological communities to anthropogenic climate and land use change, which threaten to disrupt existing patterns of species interactions and niche relationships [@Tylianakis2008-fi].

Plant-pollinator networks are a special case of resource selection in which consumer fitness and resources fitness are positively correlated, causing the traits of plants and pollinators to converge toward compatibility with one another [e.g. @Newman2014-ke]. The coexistence of multiple plant and pollinator species, however, requires niche differentiation [@Palmer2003-ih; @Johnson2019-su], which can be achieved by trait-based partitioning of interactions among pollinators and their floral hosts [@Willmer2011-ak].

Bumble bees (Hymentoptera: *Bombus* spp.) and their floral hosts provide a diverse yet tractable system in which to explore resource selection in the context of plant-pollinator networks. Due to their adaptation to cold temperatures, bumble bees are sensitive to climate change [@Goulson2008-oe; @Soroye2020-zb], especially in the mountain ecosystems that host the world's richest bumble bee communities. Warming climates have already caused measurable upslope shifts in mountain bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased co-occurrence. At the same time, socioeconomic factors have led to widespread abandonment and reforestation of mountain meadows, particularly in the European Alps, which dramatically reduces floral resource availability and suppresses bumble bee diversity and abundance [@Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw]. The ability of bumble bee species to coexist under these conditions of reshuffled distributions and declining floral resources will depend in part on how they select and partition floral resources, which will in turn affect the viability of their floral mutualists, many of which depend principally on bumble bee visitation for pollination [@Goulson2010-lo].

Addressing the question of how bumble bees will select floral resources under novel conditions requires discernment of the generalizable mechanisms that underpin patterns of floral selection observed under current conditions. It is ordinarily assumed that the choice of resources by foraging animals is a joint function of trait-based preferences and the neutral effects of the relative abundances of available resources: the so-called "use-availability" model of resource selection [e.g. @Beyer2010-vr]. The most well-studied mechanism of the preference component of bumble bees floral selection is trait-matching between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], leading to a well-documented association between tongue length and floral preferences in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. Under the use-availability model, tongue-length-based floral preferences should be modulated by the relative abundance of available flora to generate observed patterns of floral selection. It has recently been proposed, however --- in what has been termed the *target use hypothesis* --- that floral selection by bumble bees is a process of *pure preference*, irrespective of resource abundance [@De_Valpine2013-be; @Harmon-Threatt2017-bz]. If and when the *target use hypothesis* holds, observed visitation rates are a better index of preference than metrics designed to control for resource abundance (e.g. null model residuals), since the subtraction of presumed neutral effects of abundance when those effects do not really exist actually creates new abundance-dependent artifacts [@Harmon-Threatt2017-bz]. On the other hand, failing to account for neutral effects of abundance when they do exist similarly obscures the general mechanisms behind observed foraging patterns [@Vazquez2003-qa; @Bluthgen2008-rp] that will determine bumble bee resource selection under novel conditions induced by climate and land use change.

In this study, we use ~13,000 records of bumble bee visits to wildflowers, collected over three years along a 1400 m elevation gradient, to explore three unanswered questions regarding resource selection by bumble bees and its response to climate and land use change. First, we ask how bumble bee floral preferences map onto a generalized floral morphotypology [@Kugler1970-gb]. Then, using the well-established relationship between tongue length and floral selection as a point of reference, we test the *target use hypothesis* by comparing the explanatory power of bumble bee tongue length with respect to observed visitation rates vs. the residuals of a null model that assumes a linear dependency of selection on resource abundance. Under the *target use hypothesis*, tongue length should explain observed visitation rates better than null model residuals, while the opposite finding would indicate that, contrary to the *target use hypothesis*, bumble bee floral selection is influenced by relative abundance. Finally, we model the variation of bumble bee floral preference through time and elevation as a first step toward understanding how climate and land use change might distort the temporal and elevational structure of bumble- bee-wildflower mutualisms in mountain ecosystems.

@@

@@

## 3.1 Sampling overview

A total of 12,918 bumble- bee-flower interactions were included in our final data set, including 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris-lucorum* ("telu"), and *B. wurflenii* ("wurf") were the most abundant species, and they occurred at all sites. *B. hortorum* ("hort"), *B. jonellus* ("jone"),  *B. monticola* ("mont"), and *B. psithyrus* ("psit") occurred at moderate abundance. The remaining species --- *B. gerstaeckeri* ("gers"), *B. hypnorum* ("hypn"), *B. lapidarius* ("lapi"), *B. mendax* ("mend"), *B. mucidus* ("muci"), and *B. pyrenaeus* ("pyre") --- were relatively rare. *B. humilis* ("humi") was recorded only three times and only in one year, so we omitted it from all analyses.

Floral surveying yielded a total of 352 plant species (191 genera, 52 families), of which 155 (103 genera, 35 families) were observed to be visited by bumble bees. Visited floral species included all morphotypes except spike (8) and trap (10) flowers. Null (0) and brush (9) morphotypes were, however, excluded from analysis for the reasons given above.

<!-- Check again and make sure all the numbers given here are still correct. -->

@@

Preference was generally strong, with the visitation rates of most bumblebee species to most floral morphotypes falling significantly above or below the 99% interval of the null model **(Figure 1)**. Nonsignificant preference was observed mainly among the rarer bumblebee species because their small samples sizes lead to wide null model confidence intervals.

```{r fig1, echo = FALSE,  fig.cap = paste("Observed visitation rates compared to null model simulations for each bumble bee species and floral morphotype (A). Gray bars depict 99\\% intervals of null model simulations and points depict observed visitation rates. Points are color-coded according to where they fall with respect to the 99\\% interval of the null model: red = above (positive preference), blue = below (negative preference), gray = within (nonsignificant preference). The difference between the observed value and the center of the null model interval can be interpreted as magnitude of preference. Each floral morphotype is illustrated (B) with a representative species common in our study system: 1 = Potentilla erecta, 2 = Gentiana asclepiadea, 3 = Campanula glomerata, 4 = Silene nutans, 5 = Ajuga reptans, 6 = Lotus corniculatus, 7 = Centaurea jacea. Illustrations are modifications of public domain works obtained from www.plantillustrations.org. For details of original sources, see Appendix X.")}
knitr::include_graphics("../ms_2/figures/fig1.png")
```

When floral morphotypes are compared to one another in terms of preference, clear patterns emerge **(Figure 2A)**. All species exhibited negative preference toward disc flowers, while most species exhibited positive preference toward bell, lip, and flag flowers. preference toward funnel and stalk-disc flowers tended to be neutral (with one strong positive outlier --- *B. gerstaeckeri* --- for funnel flowers), reflecting the fact that these morphotypes tended both to be rarely visited and to occur at low abundance. preference toward head flowers varied widely from strongly positive to strongly negative.

Differences between floral morphotypes in terms of raw mean visitation rates, reflecting the target use hypothesis, were less pronounced **(Figure 2B)**. Bell, lip, flag, and head flowers were visited at high rates by one or more bumble bee species, but median raw mean visitation rates did not differ strongly across morphotypes.

The omission of queens from the calculation of overall preference did not yield qualitatively different results (Appendix X).

@@

<!-- Before we finalize this, check to make sure the raw visitation rates were really prepared correctly and are not being deflated by a bunch of zeros that should'nt be there. -->

```{r fig2, echo = FALSE,  fig.cap = paste("Preference (A) and observed visitation rate (B) of the bumble bee species toward each floral morphotype. Each point represents the preference/visitation of one bumble bee species toward the corresponding floral morphotype. The color of each point in panel A, retained from Figure 2, indicates whether a given observed bias is significantly (p < 0.01) positive or negative with respect to null model simulations.")}
knitr::include_graphics("../ms_2/figures/overall_ktype.png")
```

## 3.4 Tongue length as a predictor of floral choice

Tongue length was a strong predictor of preference but not of observed visitation rates **(Figure 3)**. preference toward funnel, bell, lip, flag, and head flowers exhibited significant GAM curves in response bumble bee tongue length, and the model explained 77% of total deviance (adj. R^2^ = 0.728). For funnel and lip flowers, preference increased linearly with tongue length. Preference toward bell flowers and head flowers exhibited approximately quadratic responses with preference peaking among short-tongued species and declining sharply with increasing tongue length before leveling off. preference toward flag flowers exhibited a unimodal peak among medium-tongued species. In contrast, observed visitation rates were significantly predicted by tongue length only for head flowers, which exhibited a weakly quadratic decline in preference from a peak among short-tongued species. The visitation rate model explained only 25% of total deviance (adj. R^2^ = 0.053).

Again, these results did not differ qualitatively when preference was based on the worker-only null model (Appendix X).

```{r fig3, echo = FALSE,  fig.cap = paste("preference (A) and observed visitation rates (B) for each morphotype regressed against bumble bee tongue length. Each point represents the preference/visitation of one bumble bee species toward the corresponding floral morphotype. Curves depict fitted GAM smooths with 95\\% confidence bands. Solid lines indicate curves differing significantly (p < 0.05) from a flat line while dashed lines indicated nonsigificant curves. Points are color-coded by species, and species are presented in the legend in order of increasing tongue length.")}
knitr::include_graphics("../ms_2/figures/ktypeXtonguelength_paired.png")
```

PCA of preference captured more than 80% of total variance in the first two axes **(Figure 4)**, and the GAM surface fit to the firs two axes explained most of the variance in tongue length (deviance explained = 94.7%, adj. R^2^ = 0.893, p < 0.001).

<!-- Need to add a bit of comment here about important patterns in the ordination. -->

Perhaps the most striking pattern

```{r fig4, echo = FALSE,  fig.cap = paste("Principal components biplot of bumble bee species according to their preference toward each floral morphotype. Floral morphotypes are abbreviated with their numerical codes and illustrated as in Figure 1. Colored bands depict a 2-dimensional GAM smooth representing the relationship between the first two principal components and bumble bee tongue length. Morphotype illustrations are as described in Figure 1.")}
knitr::include_graphics("../ms_2/figures/preference_ordination_mod.png")
```

## 3.5 Variation of floral abundance, raw visitation, and preference through elevation and time

The proportional abundance of the most visited floral morphotypes --- bell, lip, flag, and head flowers --- was strongly patterned in time and elevation **(Figure 5)**. Bell flowers peaked in proportional abundance early in the season and at mid-elevation, and lip flowers peaked late in the season and at low elevation. Flag flowers peaked in mid-season and exhibited two distinct elevational peaks, the lower at around 1000 m.s.l and the higher at around 1800 m.s.l. Head flowers exhibited a mid-season peak at low elevation and a broad late season peak at around 1600 m.s.l. All smooths were significant (p << 0.0001), and the model explained 64.3% of total deviance (adj. R^2^ = 0.554).

```{r fig5, echo = FALSE,  fig.cap = paste("Variation in proportional flower cover through elevation and modeled as HGAM surfaces. Color ramps depict z-normalized effects to allow the range of effect size to be compared across panels, but contour lines are set at the same fixed density across all panels to emphasize the within-panel patterns. All smooths are significant (p < 0.05).")}
knitr::include_graphics("../ms_2/figures/flower_cover_gam.png")
```

Patterns of preference through elevation and time were more complex than those of raw visitation and, by design, more often differed from patterns of floral abundance **(Figure 6)**. In some cases, though, patterns of preference mirrored patterns of floral availability, suggesting that bumble bees may exhibit functional responses [@Mysterud1998-rs] to floral abundance not entirely accounted for by null models that assume a linear proportionality between abundance and visitation.

One of the more salient patterns is that positive selection for flag flowers was strongly associated with high elevation, except among the species that tended to avoid flag flowers (*B. pratorum* and *B. soroeensis*) **(Figure 1)**. It is also apparent that, as a rule, preference is more dynamic where it is also higher overall. Bell flowers, which were positively selected overall by the short-tongued bumble bee species *B. pratorum*, "*B. terrestris-lucorum*, and *B. soroeensis* **(Figure 1)**, also exhibited the most complex patterns of preference in elevation and time for these bumble bee species. In contrast, the long and medium-tongued species *B. hortorum*, *B. pascuorum*, and *B. wurflenii*, which exhibited neutral or weakly positive overall selection toward bell flowers **(Figure 1)**, also show weak patterns of variation in preference toward bell flowers. Nearly the reverse pattern is evident for flag flowers, reflecting their association with longer-tongued bumble bee species.

Smooths were significant (p < 0.05) for all bb-fl pairs except hort-head, prat-flag, telu-head, and soro-flag. The model explained 36.4% of total deviance (adj. R^2^ = 0.33).

```{r fig6, echo = FALSE,  fig.cap = paste("Variation in preference through elevation and time modeled as HGAM surfaces. Color ramps depict z-normalized effects to allow the range of effect size to be compared across panels, but contour lines are set at the same fixed density across all panels to emphasize the within-panel patterns. Significant (p < 0.05) smooths are depicted with solid contour lines and nonsignificant smooths with dashed contour lines.")}
knitr::include_graphics("../ms_2/figures/preference_lability_gam.png")
```

@@

@@

It is well-known that bumble bees forage non-randomly accordingly to morphological constraints, innate preferences, and experienced rewards [e.g. @Kugler1943-uy; @Brian1954-cy; @Inouye1978-qi; @Raine2007-nq; @Maharaj2019-wm]. In this regard, the results of our null model analysis are as unsurprising as they are unambiguous: bumble bee resource selection almost never represents flora in unaltered proportion to their abundance. The *target use hypothesis* [@De_Valpine2013-be; @Harmon-Threatt2017-bz], however, extends this idea to the perfect antithesis of the random foraging model, positing that bumble bee foraging is a process of *pure* preference, completely irrespective of resource abundance.

Given the well-established mechanistic link between bumble bee tongue length and floral selection [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn], our finding that bumble bee tongue length explained preference (i.e. null model residuals) much better than it explained raw visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral choice. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging. It does not, however, mean that the *target use hypothesis* does not obtain under certain circumstances, such as when a highly preferred resource varies within a nonlimiting rage of availability. Moreover, our results suggest something akin to the *target use hypothesis* but formulated negatively with respect to resources that are *not selected*. Visitation rates to disc flowers in our system, for example, were very low despite the fact that they often occurred at high abundance, suggesting that while preference seems to interact with resource abundance, avoidance may not. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of abundance and the directed effect of preference, which implies that analytical techniques designed to disentangle these effects, such as the null model analysis we employ, should be favored in studies of bumble bee foraging. 
<!-- Accordingly, we shall focus primarily on our calculations of preference, rather than raw visitation rates, as we interpret the other results of our study. -->

## 4.2 Overall patterns of preference

We found that positive selection was concentrated on 3 floral morphotypes: bell, lip, and flag flowers. These results are broadly consistent with the raw visitation patterns previously reported by @Schneller2014-ry in the central Alps at Vorarlberg, Neumayer1999-il in the Austrian Alps ~55 km south of our study system, and Farnberger1996-xo in the vicinity of Vienna, suggesting that our findings are at least regionally generalizeable. These neighboring studies tended to find a stronger role of head flowers than suggested by our preference analysis, but this is probably because they characterized foraging in terms of raw visitation rather than preference. Indeed, our raw visitation data similarly indicate a stronger role of head flowers, but our preference analysis reveals that while head flowers were highly visited, they tended to be under-visited relative to their abundance.

>>>>

---
title: "Floral preferences of mountain bumble bees are constrained by trait matching but flexible through elevation and season"
header-includes:
   - \usepackage{lineno}\linenumbers
   - \usepackage{setspace}\doublespacing
output: pdf_document
bibliography: 
  - references.bib
csl: ecologyletters.csl
---

**Douglas Sponsler** (corresponding author)  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
Biozentrum der Universität Würzburg C-018  
Am Hubland  
97074 Würzburg, Germany  
Email: douglas.sponsler@uni-wuerzburg.de  
Telephone: +49 0176 363 20273

**Katharina Kallnik**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: katharina.kallnik@uni-wuerzburg.de

**Fabrice Requier**  
Université Paris-Saclay  
Évolution, Génomes, Comportement et Écologie  
91198 Gif-sur-Yvette, France  
Email: fabrice.requier@egce.cnrs-gif.fr

**Alice Classen**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: alice.classen@uni-wuerzburg.de

**A. Fabienne Maihoff**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: anne_fabienne.maihoff@uni-wuerzburg.de

**Johanna Sieger**  
Naturpark Frankenhöhe  
91522 Ansbach, Germany   
Email: sieger@lpv-mfr.de

**Ingolf Steffan-Dewenter**  
University of Würzburg  
Department of Animal Ecology and Tropical Biology  
97074 Würzburg, Germany  
Email: ingolf.steffan@uni-wuerzburg.de

\

**AUTHORSHIP STATEMENT:** KK and ISD conceived the study. KK, JS, and ISD designed and carried out field work. DS analyzed the data and wrote the paper with substantial contributions from FR, AC, AFM, KK, and ISD.

**DATA ACCESSIBILITY:** All data used in this study have been archived on DRYAD (https://datadryad.org/stash/dataset/doi:10.5061/dryad.s1rn8pk7f) and will be available upon article publication.

**ARTICLE TYPE:** Letter

**RUNNING TITLE:** Bumble bee floral preferences

**KEYWORDS:** foraging, trait matching, pollinator, *Bombus*, alpine, niche, resource selection

**SECTION LENGTHS:**  
Abstract:  
Main text:  
References:  
Figures:

\newpage

@@

Patterns of resource use by animals can clarify how ecological communities have assembled in the past, how they currently function, and how they are likely to respond to future perturbations. Using ~13,000 records of bumble- bee-wildflower interactions along a 1400 m elevation gradient, we find that the selection of floral morphotypes by bumble bees is driven both by the directed process of trait-matching and the neutral effect of floral abundance. Patterns of floral preference also vary by elevation and season, which may be an adaptation to the strong spatiotemporal turnover of mountain floral communities. The flexibility of bumble bee foraging will be tested in coming years as climate and land use change drive rapid changes in floral resource distribution in time and space, particularly in the mountain ecosystems that host the world's richest communities of bumble bees and their floral mutualists.

\newpage

@@

Resource selection and partitioning by animals are instrumental in community assembly, species coexistence, and ecosystem function [@Chesson2000-aw; @Bascompte2007-ls]. Resource selection is fundamentally constrained by phenotypic compatibility (i.e. trait-matching) between resources and consumers, but it can also be influenced by the relative availability of resources [@Johnson1980-co] and modulated by exogenous drivers such as competition [@Pianka1974-qk]. Understanding these interacting processes is key to anticipating the response of ecological communities to anthropogenic climate and land use change, which threaten to disrupt existing patterns of species interactions and niche relationships [@Tylianakis2008-fi].

\

<!-- Plant-pollinator networks are a special case of resource selection in which consumer fitness and resources fitness are positively correlated, causing the traits of plants and pollinators to converge toward compatibility with one another [e.g. @Newman2014-ke]. The coexistence of multiple plant and pollinator species, however, requires niche differentiation [@Palmer2003-ih; @Johnson2019-su], which can be achieved by trait-based partitioning of interactions among pollinators and their floral hosts [@Willmer2011-ak]. -->

Bumble bees (Hymentoptera: *Bombus* spp.) and their floral hosts provide a diverse yet tractable system in which to explore resource selection in the context of plant-pollinator networks. Due to their adaptation to cold temperatures, bumble bees are sensitive to climate change [@Goulson2008-oe; @Soroye2020-zb], especially in the mountain ecosystems that host the world's richest bumble bee communities. Warming climates have already caused measurable upslope shifts in mountain bumble bees and their floral hosts [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc], which could introduce novel competitive interactions between bumble bee species that experience increased co-occurrence. At the same time, socioeconomic factors have led to widespread abandonment and reforestation of mountain meadows, particularly in the European Alps, which reduces floral resource availability and suppresses bumble bee diversity and abundance [@Tappeiner2006-jp; @Tasser2007-qo; @Walcher2017-kw]. The ability of bumble bee species to coexist under these conditions of reshuffled distributions and declining floral resources will depend in part on how they select and partition floral resources, which will in turn affect the viability of their floral mutualists, many of which depend principally on bumble bee visitation for pollination [@Goulson2010-lo].

\

Addressing the question of how bumble bees will select floral resources under novel conditions requires discernment of the generalizable mechanisms that underpin patterns of floral selection observed under current conditions. It is ordinarily assumed that the choice of resources by foraging animals is a joint function of trait-based preferences and the neutral effects of the relative abundances of available resources: the so-called "use-availability" model of resource selection [e.g. @Beyer2010-vr]. The most well-studied mechanism of the preference component of bumble bees floral selection is trait-matching between bumble bee tongue length and floral corolla depth. Bumble bee species can differ markedly in tongue length, and floral handling time increases when tongue length and corolla depth are poorly matched [@Inouye1980-mv; @Harder1983-bk], leading to an association between tongue length and floral preferences in bumble bee communities [e.g. @Brian1954-cy; @Heinrich1976-wc; @Teras1976-pi; @Inouye1978-qi; @Pyke1982-gn]. Under the use-availability model, tongue-length-based floral preferences should be modulated by the relative abundance of available resources to generate observed patterns of floral selection. It has recently been proposed, however --- in what has been termed the *target use hypothesis* --- that floral selection by bumble bees is a process of *pure preference*, irrespective of resource abundance [@De_Valpine2013-be; @Harmon-Threatt2017-bz]. If and when the *target use hypothesis* holds, observed visitation rates are a better index of preference than metrics designed to control for resource abundance (e.g. null model residuals), since the subtraction of presumed neutral effects of abundance when those effects do not really exist actually creates new abundance-dependent artifacts [@Harmon-Threatt2017-bz]. On the other hand, failing to account for neutral effects of abundance when they do exist similarly obscures the general mechanisms behind observed foraging patterns [@Vazquez2003-qa; @Bluthgen2008-rp] that will determine bumble bee resource selection under novel conditions induced by climate and land use change.

\

In this study, we use ~13,000 records of bumble bee visits to wildflowers, collected over three years along a 1400 m elevation gradient, to explore three unanswered questions regarding resource selection by bumble bees and its response to climate and land use change. First, we ask how bumble bee floral preferences map onto a generalized floral morphotypology [@Kugler1970-gb]. Then, using the established relationship between tongue length and floral preference as a point of reference, we test the *target use hypothesis* by comparing the explanatory power of bumble bee tongue length with respect to observed visitation rates vs. the residuals of a null model that assumes a linear dependency of selection on resource abundance. Under the *target use hypothesis*, tongue length should explain observed visitation rates better than null model residuals, while the opposite finding would indicate that, contrary to the *target use hypothesis*, bumble bee floral selection is influenced by relative abundance. Finally, we model the variation of bumble bee floral preference through time and elevation as a first step toward understanding how climate and land use change might distort the temporal and elevational structure of bumble- bee-wildflower mutualisms in mountain ecosystems.

@@

\

@@

## 3.1 Bumble bee and floral communities

A total of 13,293 bumble- bee-flower interactions were included in our final data set, comprising 15 bumble bee species, 163 plant species (110 genera, 37 families), and 730 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris-lucorum* ("telu"), and *B. wurflenii* ("wurf") were the most abundant species, and they occurred at all sites. *B. hortorum* ("hort"), *B. jonellus* ("jone"),  *B. monticola* ("mont"), and *B. psithyrus* ("psit") occurred at moderate abundance. The remaining species --- *B. gerstaeckeri* ("gers"), *B. hypnorum* ("hypn"), *B. lapidarius* ("lapi"), *B. mendax* ("mend"), *B. mucidus* ("muci"), and *B. pyrenaeus* ("pyre") --- were relatively rare. *B. humilis* ("humi") was recorded only three times and only in one year, so we omitted it from all analyses. Floral surveying yielded a total of 354 plant species (194 genera, 53 families). Visited floral species included all morphotypes except spike (8) and trap (10) flowers. Null (0) and brush (9) morphotypes were, however, excluded from analysis for the reasons given above.

@@

Preference was generally strong, with the visitation rates of most bumblebee species to most floral morphotypes falling significantly above or below the 99% interval of the null model **(Figure 1)**. Nonsignificant preference was observed mainly among the rarer bumblebee species, likely because their small samples sizes lead to wide null model confidence intervals. Interestingly, bumble bees showed strong species-specific variation in their relative preferences of different floral morphotypes **(Figure 1)**. The omission of queens from the calculation of overall preference did not yield qualitatively different results (Appendix X).

\

When floral morphotypes are compared to one another in terms of preference across all bumble bee species, clear patterns emerge **(Figure 2A)**. All species exhibited negative preference toward disc flowers, while most species exhibited positive preference toward bell, lip, and flag flowers. preference toward funnel and stalk-disc flowers tended to be neutral (with one strong positive outlier --- *B. gerstaeckeri* --- for funnel flowers), reflecting the fact that these morphotypes tended both to be rarely visited and to occur at low abundance. Preference toward head flowers varied widely across bumble bee species, ranging from strongly positive to strongly negative.

\

Differences between floral morphotypes in terms of observed visitation rates, reflecting the target use hypothesis, were less pronounced **(Figure 2B)**. Bell, lip, flag, and head flowers were visited at high rates by one or more bumble bee species, but observed visitation rates did not differ strongly across morphotypes.

@@

## 3.3 Tongue length as a predictor of floral choice

Tongue length was a strong predictor of preference but not of observed visitation rates **(Figure 3)**. Preference toward funnel, bell, lip, flag, and head flowers exhibited significant GAM curves in response to bumble bee tongue length, and the model explained 77% of total deviance (adj. R^2^ = 0.728). For funnel and lip flowers, preference increased linearly with tongue length. Preference toward bell flowers and head flowers exhibited approximately quadratic responses with preference peaking among short-tongued species and declining sharply with increasing tongue length before leveling off. Preference toward flag flowers exhibited a unimodal peak among medium-tongued species. In contrast, observed visitation rates were predicted by tongue length only for head flowers, which exhibited a weak decline in preference from a peak among short-tongued species. The visitation rate model explained only 25% of total deviance (adj. R^2^ = 0.053). Again, these results did not differ when preference was based on the worker-only null model (Appendix X).

\

PCA of bumble bee floral preferences captured more than 80% of total variance in the first two axes **(Figure 4)**, and the GAM surface fit to the firs two axes explained most of the variance in tongue length (deviance explained = 94.7%, adj. R^2^ = 0.893, p < 0.001), reflecting the importance of trait matching between tongue length and floral morphotype as a mechanistic basis of preference. Longer tongue length was associated with deeper and more closed flower morphotypes (lip, funnel, flag), while shorter tongue length was associated with more shallow and open flower morphologies (disc, stalk-disc, bell, head). The strong affinity of *B. soroeensis* and *B. psithyrus* for head flowers is especially salient.

## 3.4 Variation of morphotype abundance and preference through elevation and season

The proportional abundance of the most visited floral morphotypes --- bell, lip, flag, and head flowers --- was strongly patterned by elevation and season **(Figure 5)**. Bell flowers peaked in proportional abundance early in the season and at mid-elevation, and lip flowers peaked late in the season and at low elevation. Flag flowers peaked in mid-season and exhibited two distinct elevational peaks, the lower at around 1000 m.s.l and the higher at around 1800 m.s.l. Head flowers exhibited a mid-season peak at low elevation and a broad late season peak at around 1600 m.s.l. All smooths were significant (p << 0.0001), and the model explained 64.3% of total deviance (adj. R^2^ = 0.554).

\

The morphotype preferences of bumble bee species were also strongly patterned through elevation and season. In some cases, though, patterns of preference mirrored patterns of floral availability, suggesting that bumble bees may exhibit functional responses [@Mysterud1998-rs] to floral abundance not entirely accounted for by null models that assume a linear proportionality between abundance and visitation. Preference for flag flowers was strongly associated with high elevation, except among the species that tended to avoid flag flowers (*B. pratorum* and *B. soroeensis*) **(Figure 1)**. It is also apparent that, as a rule, preference is more dynamic where it is also higher overall. Bell flowers, which were positively selected overall by the short-tongued bumble bee species *B. pratorum*, "*B. terrestris-lucorum*, and *B. soroeensis* **(Figure 1)**, also exhibited the most complex patterns of preference in elevation and time for these bumble bee species **(Figure 6)**. In contrast, the long and medium-tongued species *B. hortorum*, *B. pascuorum*, and *B. wurflenii*, which exhibited neutral or weakly positive overall selection toward bell flowers **(Figure 1)**, also show weak patterns of variation in preference toward bell flowers **(Figure 6)**. Nearly the reverse pattern is evident for flag flowers, reflecting their association with longer-tongued bumble bee species.

\

Smooths were significant (p < 0.05) for all bee-morphotype pairs except hort-head, prat-flag, telu-head, and soro-flag. The model explained 36.4% of total deviance (adj. R^2^ = 0.33).

@@

Using ~13,000 bumble- bee-wildflower interactions, we characterized bumble bee floral selection in terms of generalized floral morphotypes and found evidence that bumble bee floral selection depends both on the directed process of morphotype preference and the neutral effect of floral abundance. Bumble bee species differed in their floral morphotype preferences, and these differences could be explained by trait-matching between floral morphotype and bumble bee tongue length. These preferences, however, varied by elevation and season, likely representing adaptive responses of bumble bee foraging behavior to elevational and seasonal shifts in floral resource composition, possibly modulated by competitive interactions among bumble bees species with overlapping resource preferences.

@@

Our finding that bumble bee tongue length explained preference (i.e. null model residuals) much better than it explained observed visitation rates indicates that, in our study system, abundance and preference function jointly in bumble bee floral selection. This casts doubt on the *target use hypothesis* as a general rule of bumble bee foraging, though it is possible that it still obtains under certain circumstances [e.g. @Harmon-Threatt2017-bz]. In general, though, our findings support the hypothesis that bumble bee resource selection involves both the neutral effects of floral abundance and the directed effect of preference, meaning that analytical techniques designed to disentangle these effects, such as the null model method we employ, should be favored in studies of bumble bee foraging.

## 4.2 Community-level patterns of morphotype preferences

We found that the most preferred morphotypes were bell, lip, and flag flowers, which is broadly consistent with visitation patterns reported by previous studies in central Europe [@Schneller2014-ry; @Neumayer1999-il; Farnberger1996-xo], indicating that our findings are regionally generalizeable. These previous studies tended to find a stronger role of head flowers than indicated by our preference analysis, but this is probably because they characterized foraging in terms of simple visitation rates rather than null-model-inferred preference. We saw similarly high visitation rates toward head flowers, but our null-model analysis reveals that while head flowers were highly visited, they tended to be under-visited relative to their abundance.

\

While bell, lip, and flag flowers were the most preferred morphotypes on average, preference toward these morphotypes varied broadly across bumble bee species. Moreover, while average preference toward funnel and head flowers was low, each was strongly preferred by one or two bumble bee species. The majority of variation in preference toward floral morphotypes was explainable by bumble bee tongue length, with the more open bell and head morphotypes selected more positively by short-tongued bumble bee species and the deeper funnel, lip, and flag morphotypes selected more positively by medium- and long-tongued bumble bee species. The failure of tongue length to explain preference toward disc and stalk-disc flowers reflects the fact that these flowers were almost unanimously ignored by all bumble bee species. We cannot say from our data why disc and stalk-disc flowers received such consistently low visitation, but the pattern may be driven in part by the fact that some disc flowers offer pollen but little or no nectar; indeed, Kugler's (1970) typology recognizes this distinction with subdivisions of the disc morphotype into those with (class 1.2) and without (class 1.1) nectar rewards. A limitation of our study is that we could not reliably distinguish pollen- and nectar-foraging bumble bees, and it is likely that this distinction would shed further light on the mechanisms underlying the selection patterns we observed.

\
